---
title: "Sales Forecasting Web Application"
author: Yash Sankhala
date: July 15, 2023
output: html_document
runtime: shiny
---

#### Building Interactive Business Reports with R
#### DAT-3535

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(forecast)
library(plotly)
library(scales)

# historical records from multiple stores
fsales <- "https://raw.githubusercontent.com/multidis/hult-inter-bus-reports-r/main/forecasting/sales_weekly.csv"
sales <- read_csv(fsales)

# latest (current) week
nweek_now <- max(sales$Week)
```

## Exploring Weekly Sales in Different Stores

Gain insights into the variations and trends of weekly sales across different stores and make informed decisions for better business strategies.

```{r echo = FALSE}
inputPanel(
  selectInput("storenum", label = "Select store: ",
              choices = 1:max(sales$Store), selected = 25),
  sliderInput("horizon", label = "Forecast horizon (weeks): ",
              min = 2, max = 52, value = 20, step = 1),
  numericInput("week", "Select week:", min = 2, max = max(sales$Week), 
                   value = 20, step = 1),
  sliderInput("confidence_interval", label = "Confidence Interval:",
              min = 0, max = 0.95, value = 0.95, step = 0.05),
  selectInput("selected_week", "Select an upcoming week:",
                choices = c(143:(nweek_now + 12)))
)
```

## Exploring Trends, Forecasts, and Comparisons for Informed Decision-Making

The Sales Analysis Dashboard provides comprehensive insights into sales trends, accurate forecasts, and meaningful comparisons across different metrics. This interactive dashboard will empower decision-makers with the necessary information to make informed business decisions and drive success.

### Weekly Sales Trend for Stores

The graph displays the weekly sales trend for a particular store selected in the UI given above, providing insights into the overall sales patterns and fluctuations over time. 

```{r echo = FALSE}
# Plot showing actual sales
renderPlotly({
  filtered_data <- sales %>% filter(Store == input$storenum)

  sales_weekly <- filtered_data %>% 
  ggplot(aes(Week, Weekly_Sales)) +
  geom_line(color = "steelblue", size = 1) +  # Reduce line size to 1.5
  labs(x = "Weeks", y = "Weekly Sales", title = paste0("Weekly Sales Trend for Store ", input$storenum)) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10, color = "black"),  # Set axis text color to black
    axis.line = element_line(color = "black"),  # Set axis line color to black
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()  # Remove minor gridlines
  ) +
  scale_y_continuous(labels = function(x) {
    if (max(filtered_data$Weekly_Sales) >= 1e6) {
      format(x/1e6, big.mark = ",", decimal.mark = ".", scientific = FALSE) %>% paste0("M")
    } else {
      format(x/1e3, big.mark = ",", decimal.mark = ".", scientific = FALSE) %>% paste0("K")
    }
  })
})
```

### Weekly Sales Forecast for Store

The graph shows the weekly sales forecast for a selected store, providing an estimate of future sales trends based on historical data and forecasting techniques. It helps in anticipating and planning for future sales performance. By using the UI the forecast can be seen for as long as 52 weeks.

```{r echo = FALSE}
# Plot showing forecast
renderPlotly({
  horiz <- input$horizon
  
  sales_ts <- ts(sales %>% filter(Store == input$storenum) %>%
                   select(Weekly_Sales), frequency = 52)
  arima_model <- auto.arima(sales_ts, seasonal.test = "seas")
  arima_pred <- forecast(arima_model, h = horiz)
  
  sales_pred <- data.frame(Sales_Predicted = as.numeric(arima_pred$mean),
                           Week = (nweek_now + 1):(nweek_now + horiz))
  
  sales_pred %>% 
  ggplot(aes(Week, Sales_Predicted)) +
  geom_line(color = "steelblue", size = 1, linetype = "dashed") +
  labs(x = "Weeks ahead", y = "Weekly Sales Forecast", title = paste0("Weekly Sales Forecast for Store ", input$storenum)) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10, color = "black"),
    axis.line = element_line(color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_y_continuous(labels = function(x) {
    if (max(sales_pred$Sales_Predicted) >= 1e6) {
      format(x/1e6, big.mark = ",", decimal.mark = ".", scientific = FALSE) %>% paste0("M")
    } else {
      format(x/1e3, big.mark = ",", decimal.mark = ".", scientific = FALSE) %>% paste0("K")
    }
  })
})
```

### Sales Forecast for Selected Week

The graph shows the sales forecast for a specific week within the next three months, as selected in the UI. It provides an estimate of the expected sales for the chosen week, helping in understanding the projected sales performance and assisting in making informed business decisions. The forecast takes into account historical sales data and utilizes forecasting techniques to provide a reliable estimate for the selected week.

```{r echo = FALSE}
# Calculating the forecasted sales for any week upcoming within next 3 months
renderUI({
  selected_week <- as.numeric(input$selected_week)
  
  # Forecasting
  sales_ts <- ts(sales %>% filter(Store == input$storenum) %>%
                   select(Weekly_Sales), frequency = 52)
  arima_model <- auto.arima(sales_ts, seasonal.test = "seas")
  arima_pred <- forecast(arima_model, h = selected_week - nweek_now)
  sales_forecast <- arima_pred$mean[selected_week - nweek_now]
  
  # Format forecasted sales as dollar amount
  formatted_sales <- format(round(sales_forecast, digits = 2), big.mark = ",", decimal.mark = ".", scientific = FALSE)

# Determine whether to display in thousands or millions
if (sales_forecast >= 1000000) {
  formatted_sales <- paste0("$", format(round(sales_forecast/1000000, digits = 2), big.mark = ",", decimal.mark = ".", scientific = FALSE), "M")
} else {
  formatted_sales <- paste0("$", format(round(sales_forecast/1000, digits = 2), big.mark = ",", decimal.mark = ".", scientific = FALSE), "K")
}

week_label <- paste0("Week: ", selected_week)
store_number <- paste0("Store: ", input$storenum)

formatted_sales_html <- paste0("<span style='font-size: 80px; font-weight: bold; color: steelblue;'>", formatted_sales, "</span>")
week_label_html <- paste0("<span style='font-size: 30px;'>", week_label, "</span>")
store_number_html <- paste0("<span style='font-size: 30px;'>", store_number, "</span>")

div(
  style = "display: flex; justify-content: center; align-items: center; height: 100vh; flex-direction: column;",
  HTML(store_number_html),
  HTML(week_label_html),
  HTML(formatted_sales_html)
)
})
```

### 90-Day Sales Forecast from Selected Week"

The graph illustrates a 90-day sales forecast starting from the selected week. It provides valuable insights into the projected sales trends over the next three months, aiding in decision-making and planning for the business.

```{r echo = FALSE}
# Plot comparing weekly sales for the past year (actual) and upcoming year (forecast)
renderPlotly({
  # Filter sales data for the selected store
  filtered_sales <- sales %>% filter(Store == input$storenum)
  
  # Get the selected week and year
  selected_week <- input$week
  
  # Calculate the selected year
  selected_year <- floor(selected_week / 52)
  
  # Get the start and end weeks for the actual and predicted data
  start_week_actual <- (selected_year - 1) * 52 + 1
  end_week_actual <- selected_week
  start_week_predicted <- selected_week + 1
  end_week_predicted <- selected_week + (52 * 1)
  
  # Filter sales data for the actual period
  actual_sales <- filtered_sales %>%
    filter(Week >= start_week_actual, Week <= end_week_actual) %>%
    mutate(Year = "Actual")
  
  # Generate sales forecast for the predicted period
  sales_ts <- ts(filtered_sales$Weekly_Sales, frequency = 52)
  arima_model <- auto.arima(sales_ts, seasonal.test = "seas")
  arima_pred <- forecast(arima_model, h = 52 * 1)
  
  predicted_sales <- data.frame(Weekly_Sales = as.numeric(arima_pred$mean),
                                Week = (end_week_actual + 1):(end_week_predicted),
                                Year = "Predicted")
  
  # Combine the actual and predicted sales data
  combined_sales <- bind_rows(actual_sales, predicted_sales)
  
  # Plot the sales comparison
  combined_sales %>%
  ggplot(aes(x = Week, y = Weekly_Sales, color = Year, group = Year)) +
  geom_line(linetype = "solid", size = 1) +
  geom_line(data = combined_sales[combined_sales$Year == "Predicted", ], linetype = "dashed", size = 1, alpha = 0.7) +
  labs(x = "Week", y = "Sales", title = paste("Sales Forecast after Week", input$week)) +
  scale_color_manual(values = c("Actual" = "steelblue", "Predicted" = "steelblue")) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14),
    axis.line = element_line(color = "black"),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  ) +
  scale_y_continuous(labels = function(x) {
    if (max(combined_sales$Weekly_Sales) >= 1e6) {
      format(x / 1e6, big.mark = ",", decimal.mark = ".", scientific = FALSE) %>% paste0("M")
    } else {
      format(x / 1e3, big.mark = ",", decimal.mark = ".", scientific = FALSE) %>% paste0("K")
    }
  })

})
```

### Current Week Performance: Actual vs. Forecast

The graph displays the comparison between the actual sales for the current week and the forecasted sales for the same week, computed at the end of the last week. It provides an indication of the performance of sales for the current week, allowing for insights into the accuracy of the forecast and potential deviations from the expected sales.

```{r echo = FALSE}
# Plot comparing actual weekly sales and forecasted weekly sales for the selected week and store
renderPlotly({
  # Filter sales data for the selected store and week
  filtered_sales <- sales %>%
    filter(Store == input$storenum)
  
  # Get the actual sales for the selected week
  actual_sales <- filtered_sales$Weekly_Sales[filtered_sales$Week == input$week]
  
  # Use all previous data to generate sales forecast for the selected week
  sales_ts <- ts(filtered_sales$Weekly_Sales[filtered_sales$Week < input$week], frequency = 52)
  
  # Check if there are enough observations for forecasting
  if (length(sales_ts) >= 2 || input$week == 0) {
    if (input$week > 0) {
      arima_model <- auto.arima(sales_ts)
      forecasted_sales <- forecast(arima_model, h = 1)$mean
    } else {
      forecasted_sales <- 0
    }
    
    # Create a data frame for the actual and forecasted sales
    sales_data <- data.frame(Type = c("Actual", "Forecast"),
                             Sales = c(actual_sales, forecasted_sales))
    
    # Plot the actual and forecasted sales for the selected week and store as a bar graph
    plot_ly(sales_data, x = ~Type, y = ~Sales, type = "bar", color = ~Type) %>%
    layout(
      title = paste("Sales Comparison for Week", input$week, "Store", input$storenum),
      xaxis = list(
        title = "Type",
        color = "black",
        showgrid = FALSE,
        zeroline = FALSE,
        showticklabels = TRUE,
        linecolor = "black"
      ),
      yaxis = list(
        title = "Sales",
        color = "black",
        showgrid = FALSE,
        zeroline = FALSE,
        showticklabels = TRUE,
        linecolor = "black",
        tickformat = ".3s"  # Format y-axis labels in thousands or millions
      )
    )
} else {
  plot_ly() %>%
    layout(
      xaxis = list(
        title = "Type",
        showgrid = FALSE,
        zeroline = FALSE,
        showticklabels = FALSE,
        color = "black",
        linecolor = "black"
      ),
      yaxis = list(
        title = "Sales",
        showgrid = FALSE,
        zeroline = FALSE,
        showticklabels = FALSE,
        color = "black",
        linecolor = "black"
      ),
      annotations = list(
        x = 0.5,
        y = 0.5,
        text = "Insufficient data for forecasting. Increase weeks to see comparison.",
        showarrow = FALSE,
        font = list(color = "red", size = 18)
      )
    )
  }
})
```

### Forecast with Confidence Intervals: Upcoming Month

The graph displays the forecasted sales for the upcoming month, along with the selected confidence intervals. The confidence intervals provide a range of values within which the actual sales are expected to fall with a certain level of confidence. This information helps in assessing the uncertainty associated with the forecasted sales and provides insights into the potential variability in the sales performance for the upcoming month.

```{r echo = FALSE}
renderPlotly({
  horiz <- input$horizon
  
  sales_ts <- ts(sales %>% filter(Store == input$storenum) %>%
                   select(Weekly_Sales), frequency = 52)
  arima_model <- auto.arima(sales_ts, seasonal.test = "seas")
  arima_pred <- forecast(arima_model, h = horiz, level = input$confidence_interval)
  
  sales_pred <- data.frame(
    Sales_Predicted = as.numeric(arima_pred$mean),
    Lower_Bound = as.numeric(arima_pred$lower),
    Upper_Bound = as.numeric(arima_pred$upper),
    Week = (nweek_now + 1):(nweek_now + horiz)
  )
  
  # Plot forecasted sales with confidence intervals
  # Format y-axis labels as thousands or millions
  formatted_y_labels <- function(x) {
    ifelse(
      abs(x) >= 1e6,
        paste0(signif(x / 1e6, digits = 2), "M"),
    ifelse(
      abs(x) >= 1e3,
      paste0(signif(x / 1e3, digits = 2), "K"),
      as.character(x)
    )
  )
}

# Create the plot
p <- sales_pred %>% 
  ggplot(aes(Week, Sales_Predicted)) +
  geom_line(color = "steelblue", size = 1, linetype = "dashed") +
  labs(x = "Weeks", y = "Weekly Sales Forecast", 
       title = paste0("Weekly Sales Forecast with Confidence Intervals for Store ", input$storenum )) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10, color = "black"),
    axis.line = element_line(color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_y_continuous(labels = formatted_y_labels)  # Use custom y-axis labels

# Add confidence intervals to the plot
if (input$confidence_interval > 0) {
  p <- p + geom_ribbon(aes(ymin = Lower_Bound, ymax = Upper_Bound), fill = "lightblue", alpha = 0.2)
}
p
})
```

### Monthly Sales Comparison: Past Year vs. Upcoming Year

The graph compares the monthly sales between the past year (actual) and the upcoming year (forecast). It provides insights into the expected sales trends for each month in the upcoming year, allowing for a visual comparison with the sales data from the corresponding months in the previous year. This comparison helps in identifying any potential patterns, seasonality, or changes in sales performance, aiding in decision-making and strategic planning.

```{r echo = FALSE}
renderPlotly({
  selected_store <- input$storenum
  nweek_now <- max(sales$Week)
  
  # Filter sales data for the selected store
  store_sales <- sales %>%
    filter(Store == selected_store)
  
  # Calculate the monthly sales
  store_sales_monthly <- store_sales %>%
    mutate(Month = floor((Week - 1) / 4) + 1) %>%
    group_by(Month) %>%
    summarize(Weekly_Sales = sum(Weekly_Sales))
  
  # Calculate the last 12 months of actual sales
  last_12_months_sales <- store_sales_monthly %>%
    filter(Month > (max(Month) - 12))
  
  # Create a time series object
  sales_ts <- ts(store_sales_monthly$Weekly_Sales, frequency = 12)
  
  # Build ARIMA model and forecast for the next 12 months
  arima_model <- auto.arima(sales_ts, seasonal.test = "seas")
  arima_pred <- forecast(arima_model, h = 12)
  
  # Generate forecasted sales data
  sales_pred <- data.frame(
    Sales_Predicted = as.numeric(arima_pred$mean),
    Month = (max(store_sales_monthly$Month) + 1):(max(store_sales_monthly$Month) + 12)
  )
  
  # Create an interactive plot using plot_ly
  plot_ly() %>%
    add_lines(data = last_12_months_sales, x = ~Month, y = ~Weekly_Sales, color = I("steelblue"), name = "Actual",
              line = list(dash = "solid", width = 3, alpha = 1)) %>%
    add_lines(data = sales_pred, x = ~Month, y = ~Sales_Predicted, color = I("steelblue"), name = "Forecast",
              line = list(dash = "dash", width = 3, alpha = 0.7)) %>%
    layout(
      xaxis = list(title = "Month", color = "black", linewidth = 1, showgrid = FALSE),
      yaxis = list(title = "Monthly Sales", color = "black", linewidth = 1, tickformat = ",.3s", showgrid = FALSE),
      title = list(
        text = paste0("Actual and Forecast Comparison of Monthly Sales for a Year for Store ", input$storenum),
        font = list(color = "black", size = 16)
      ),
      plot_bgcolor = "white",
      paper_bgcolor = "white",
      theme = "minimal"
    )
})
```